---
title: 'Datathon Presentation'
author: "Carolyn Chen, José San Martin, Michael Tan, Man-Lin Hsiao"
date: "11/7/2018"
output:
  beamer_presentation:
    includes:
      in_header: custom2.tex
  ioslides_presentation: default
font-size: 12px
---
```{r, echo=FALSE}
library(rmarkdown)
library(knitr)
suppressWarnings(suppressMessages(library(data.table)))
suppressWarnings(suppressMessages(library(plyr)))
suppressWarnings(suppressMessages(library(dplyr)))
suppressWarnings(suppressMessages(library(tidyr)))
suppressWarnings(suppressMessages(library(vioplot)))
suppressWarnings(suppressMessages(library(ggplot2)))

user_profile = fread("user_profile.csv")
first_session = fread("first_session.csv")
user_engagement = fread("user_engagement.csv")
```

Exploratory Data Analysis
===

- Credit Sesame is a credit and loan-management platform
- Datasets: User Profile, First Session, 30-Day User Engagement
- First, we wanted to understand the demographics of Credit Sesame users. 
- Data cleaning for ease of visualization
- Histograms, Dot Plots, Violin Plots, Choropleth Maps
```{r, echo=FALSE, out.width='.49\\linewidth', fig.align='center', fig.width=5, fig.height=5}

#Barplot of Credit Score Buckets
barplot(table(user_profile$credit_score_bucket), main = "Credit Score Frequency", cex.names=0.5, cex.main=0.8)

#make age and credit score continuous
sub<-function(x){
 x <-x%>%
   gsub("[(]", "",.)%>%
   gsub("[]]", "",.)%>%
   gsub(" ", "", .)%>%
   strsplit(split = ",")
 for(i in 1:length(x)){
   x[[i]] = as.numeric(x[[i]][1]) + as.numeric(x[[i]][2])
   x[[i]] = x[[i]]/2
   }
   return(unlist(x))
}
user_profile$credit_score_continuous = sub(user_profile$credit_score_bucket)
user_profile$age_continuous = sub(user_profile$age_bucket)

```

Exploratory Data Analysis (cont.)
===

- Dot plots of engagement stats versus credit score
```{r,echo=FALSE}
c = merge(user_profile, user_engagement, on="user_id", how="outer")

#Types of loans by gender
total_open_loan_accounts = list(user_profile$count_tradelines_open_mortgages, user_profile$count_tradelines_open_student_loans, user_profile$count_tradelines_open_secured_loans, user_profile$count_tradelines_open_unsecured_loans)

par(mfrow=c(1,2))
count1 = 1
engagementStats = list(c$view_count, c$click_count, c$click_apply_count, c$session_length)
engagementNames = c("view count", "click count", "click apply count", "session length")
y = c("count", "count", "count", "time (seconds)")
for(count in engagementStats){
  if(count1==3|count1==4){
    plot(c$credit_score_continuous, count, main = c(engagementNames[count1]), xlab = "credit score", ylab=y[count1])
  }
  count1=count1+1
}

```

EDA Visualizations (cont.)
===

- Violin plots of loan type vs homeowners
```{r, echo=FALSE}

count = 1
tmp = c("mortgages", "student loans", "secured loans", "unsecured loans")
par(mfrow=c(1,2))
for(numAccounts in total_open_loan_accounts){
  if(count==1 | count ==3) {
      vioplot(numAccounts[user_profile$is_homeowner==TRUE], numAccounts[user_profile$is_homeowner==FALSE], names = c("T", "F"))
  title(main=tmp[count], ylab = "count", xlab="homeowner")
  }
  count=count+1
}

```


EDA Visualizations (cont.)
===

- Violin Plots of loan type vs gender
```{r,echo=FALSE}
count = 1
tmp = c("mortgages", "student loans", "secured loans", "unsecured loans")
par(mfrow=c(1,2))
for(numAccounts in total_open_loan_accounts){
  if(count==1 | count ==3){
     vioplot(numAccounts[user_profile$gender==""], numAccounts[user_profile$gender=="Female"], numAccounts[user_profile$gender=="Male"], numAccounts[user_profile$gender=="Unisex"], names = c("Unknown", "F", "M", "Unisex"))
  title(main=tmp[count], ylab = "count", xlab="gender")
  }
   count=count+1
 
}
```


EDA Visualizations (cont.)
===

- Choropleth Map shows us geographical distribution of credit card debt trends
- Delinquency: user has missed 2 consecutive payments
- What are profiles of delinquent vs. non-delinquent users and within levels of delinquency?
```{r, echo=FALSE,fig.width=4, fig.height=4,fig.align='center'}
suppressWarnings(suppressMessages(library(devtools)))
suppressWarnings(suppressMessages(library(tidyverse)))
suppressWarnings(suppressMessages(library(usmap)))
#usmap::plot_usmap()

AvgTotalCCAmountPastDue = user_profile %>% group_by(state) %>% summarize(mean(total_open_cc_amount_past_due))


plot_usmap(data = AvgTotalCCAmountPastDue , values = "mean(total_open_cc_amount_past_due)", lines = "brown") + 
  scale_fill_continuous(
    low = "white", high = "brown", name = "Avg CC Amount Past Due", label = scales::comma
  ) + theme(legend.position = "right")
```