---
title: 'Understanding Credit Card Delinquency'
author: "Carolyn Chen, José San Martin, Michael Tan, Man-Lin Hsiao"
date: "11/7/2018"
output:
  beamer_presentation:
    includes:
      in_header: custom2.tex
  ioslides_presentation: default
font-size: 12px
---
```{r, echo=FALSE}
suppressWarnings(suppressMessages(library(data.table)))
suppressWarnings(suppressMessages(library(dplyr)))
suppressWarnings(suppressMessages(library(plyr)))
suppressWarnings(suppressMessages(library(tidyr)))
suppressWarnings(suppressMessages(library(vioplot)))
suppressWarnings(suppressMessages(library(ggplot2)))
suppressWarnings(suppressMessages(library(magrittr)))
suppressWarnings(suppressMessages(library(devtools)))
suppressWarnings(suppressMessages(library(tidyverse)))
suppressWarnings(suppressMessages(library(usmap)))
suppressWarnings(suppressMessages(library(randomForest)))
suppressWarnings(suppressMessages(library("cluster")))
suppressWarnings(suppressMessages(library("factoextra")))

user_profile = fread("user_profile.csv")
first_session = fread("first_session.csv")
user_engagement = fread("user_engagement.csv")
```

Data Set Visualizations
===

- Credit Sesame is a credit and loan-management platform
- Datasets: User Profile, First Session, 30-Day User Engagement
- First, we wanted to understand the demographics of Credit Sesame users. 
- Data cleaning for ease of visualization
- Histograms, Dot Plots, Violin Plots, Choropleth Maps
```{r, echo=FALSE, out.width='.49\\linewidth', fig.width=3, fig.height=3,fig.align='center'}
#Barplot of Credit Score Buckets
barplot(table(user_profile$credit_score_bucket), main = "Credit Score Frequency")
#make age and credit score continuous
sub<-function(x){
 x <-x%>%
   gsub("[(]", "",.)%>%
   gsub("[]]", "",.)%>%
   gsub(" ", "", .)%>%
   strsplit(split = ",")
 for(i in 1:length(x)){
   x[[i]] = as.numeric(x[[i]][1]) + as.numeric(x[[i]][2])
   x[[i]] = x[[i]]/2
   }
   return(unlist(x))
}
user_profile$credit_score_continuous = sub(user_profile$credit_score_bucket)
user_profile$age_continuous = sub(user_profile$age_bucket)
```


Exploratory Data Analysis (cont.)
===

- Dot plots of engagement stats versus credit score
```{r,echo=FALSE}
c = merge(user_profile, user_engagement, on="user_id", how="outer")
#Types of loans by gender
total_open_loan_accounts = list(user_profile$count_tradelines_open_mortgages, user_profile$count_tradelines_open_student_loans, user_profile$count_tradelines_open_secured_loans, user_profile$count_tradelines_open_unsecured_loans)
par(mfrow=c(1,2))
count1 = 1
engagementStats = list(c$click_apply_count, c$session_length)
engagementNames = c("click apply count", "session length")
y = c("count", "time (seconds)")
for(count in engagementStats){
    plot(c$credit_score_continuous, count, main = c(engagementNames[count1]), xlab = "credit score", ylab=y[count1], pch='.', col=rgb(0,0,0,alpha=0.5))
  count1=count1+1
}
```

EDA Visualizations (cont.)
===
- Violin Plots of loan type vs gender
```{r,echo=FALSE}
count = 1
tmp = c("mortgages", "student loans", "secured loans", "unsecured loans")
par(mfrow=c(1,2))
for(numAccounts in total_open_loan_accounts){
  if(count==1 | count ==3){
     vioplot(numAccounts[user_profile$gender==""], numAccounts[user_profile$gender=="Female"], numAccounts[user_profile$gender=="Male"], numAccounts[user_profile$gender=="Unisex"], names = c("Unknown", "F", "M", "Unisex"))
  title(main=tmp[count], ylab = "count", xlab="gender")
  }
   count=count+1
 
}
```

EDA Visualizations (cont.)
===
- Violin plots of loan type vs homeowners
```{r, echo=FALSE}
count = 1
tmp = c("mortgages", "student loans", "secured loans", "unsecured loans")
par(mfrow=c(1,2))
for(numAccounts in total_open_loan_accounts){
  if(count==1 | count ==3) {
      vioplot(numAccounts[user_profile$is_homeowner==TRUE], numAccounts[user_profile$is_homeowner==FALSE], names = c("T", "F"))
  title(main=tmp[count], ylab = "count", xlab="homeowner")
  }
  count=count+1
}
```


Visualizations (cont.)
===

- Choropleth Map shows us geographical distribution of credit card debt trends
- Delinquency: user has missed 2 consecutive payments
- What are profiles of delinquent vs. non-delinquent users and within levels of delinquency?
```{r, echo=FALSE,fig.width=5, fig.height=4,fig.align='center'}
#usmap::plot_usmap()
AvgTotalCCAmountPastDue = user_profile %>% dplyr::group_by(state) %>% dplyr::summarize(mean(total_open_cc_amount_past_due))
plot_usmap(data = AvgTotalCCAmountPastDue , values = "mean(total_open_cc_amount_past_due)", lines = "brown") + 
  scale_fill_continuous(
    low = "white", high = "brown", name = "Avg CC Amount Past Due", label = scales::comma
  ) + theme(legend.position = "right")
```

Next Step: Random Forest Model
===
- From our EDA we could already tell that there was a difference in profile between people with and without derogatory accounts, and the next logical step would be to create a predictive model
- Appropriate model given we have response variable for derogratory variable

```{r, echo=FALSE, results= "hide"}
#Set Up Data 
RFData <- user_profile %>%
  mutate(derogatory = ifelse(count_tradelines_condition_derogatory==0, 0, 1))%>%
  dplyr::select(-c(user_signup_timestamp,user_id, recent_bankruptcy_date,age_bucket, credit_score_bucket, gender, zipcode, state, count_tradelines_condition_derogatory))%>%
  dplyr::select(-c(1))%>%
  na.omit()
RFData$derogatory = as.factor(RFData$derogatory)
#Set up training and testing data
set.seed(123)
smp_size <- floor(0.75 * nrow(RFData))
train_ind <- sample(seq_len(nrow(RFData)), size = smp_size)
train <- RFData[train_ind, ]
test <- RFData[-train_ind, ]
#Train the model
Derog.rf=randomForest(derogatory ~ . , data = train)
#Obtain Accuracy
pred_test = predict(Derog.rf, test, type="response")
mean(pred_test == test$derogatory)#About a 90% accuracy!
```

Random Forest Model 
===
- Accuracy of about 90%, obtained a pretty small out-of-bag error rate
```{r,warning=FALSE, message=FALSE, echo=FALSE, fig.width=4, fig.height=3,fig.align='center'}
#Look at the OOB error vs number of trees
plot(Derog.rf)
```

Random Forest Model
===
- Now we look at other values from our fitted Random Forest model
- Using a classifier to determine if a person has obtained a derogatory account, or not at all. We concluded that having even just one derogatory account is cause of concern for the bank. 
- Based on the importance output, the 5 variables of the highest importance are homeownership, tradelines average days since opened, tradelines maximum days since opened, tradelines minimum days since opened, and number of closed tradeline accounts 

Poisson Model 
===
\begin{equation*}
  \begin{aligned}
    \mathrm{Derogratory Tradelines} = \alpha 
    &+ \beta_{1}  \mathrm{Age} \\
    &+ \beta_{2}  \mathrm{Gender} \\
    &+ \beta_{3}  \mathrm{Credit Score} \\
    &+ \beta_{4}  \mathrm{Credit Card Utilization Ratio} \\
    &+ \beta_{5}  \mathrm{Auto Loans Balance} \\
    &+ \beta_{6}  \mathrm{Student Loans Balance} \\
    &+ \beta_{7}  \mathrm{Mortgage Balance}  \\
    &+ \beta_{8}  \mathrm{Mortgage Loan * Auto Loan} \\
    &+ \beta_{9}  \mathrm{Auto Loan * Student Loan}
  \end{aligned}
\end{equation*}

\smallbreak
\begin{small}
- Age: For every 10 years we add to a user, we expect the number of derogatory accounts to change by a multiplicative factor of e(10*9.863e-03) = 1.1. This shows age is not a significant factor.
\smallbreak
When we increase the credit score of a user by 100 points, the expected number of derogatory accounts changes by a multiplicative factor of 0.36.
\end{small}
```{r,echo=FALSE, warning=FALSE}
#Poisson Model
pois_data <- user_profile%>%
 filter(gender != "")
#Poisson Model
pois_data <- user_profile%>%
 filter(gender != "")
creditScore1 = sub(pois_data$credit_score_bucket)
Age1 = sub(pois_data$age_bucket)
m1 <- glm(count_tradelines_condition_derogatory ~ Age1 + gender + creditScore1 + avg_cc_utilization_ratio + total_auto_loans_balance + total_student_loans_balance + total_mortgage_loans_balance + total_auto_loans_balance:total_student_loans_balance + total_auto_loans_balance:total_mortgage_loans_balance, family="poisson", data=pois_data)
```

Conclusions
===

- Users with non-zero derogatory accounts tended to have similar characteristics regardless of how many of those accounts they had. Once one deliquency noted, intervention should occur to stem further ones.
- Age should not be a significant predictor of user derogatory behavior. 
- Credit score is a strong predictor of derogatory behaviour, but user base of CS is also right-skewed.
- States with most past due credit card accounts are Nevada, Utah, Wyoming and Vermont. Could target reminders to people from those states.


Original Model: Clustering 
===
- K means clustering to identify similarity of deliquent vs non-deliquent users
- Looked at users with 'none' (0), 'some' (1-2) and 'many' (>2) deliquencies
- Users with 'some' deliquencies still quite similar to those with 'many'

```{r, echo=FALSE, fig.width=4, fig.height=3,fig.align='center'}
#make age and credit score continuous
sub<-function(x){
 x <-x%>%
   gsub("[(]", "",.)%>%
   gsub("[]]", "",.)%>%
   gsub(" ", "", .)%>%
   strsplit(split = ",")
 for(i in 1:length(x)){
   x[[i]] = as.numeric(x[[i]][1]) + as.numeric(x[[i]][2])
   x[[i]] = x[[i]]/2
   }
   return(unlist(x))
}
user_profile$credit_score_continuous = sub(user_profile$credit_score_bucket)
user_profile$age_continuous = sub(user_profile$age_bucket)
c = merge(user_profile, user_engagement, on="user_id", how="outer")

cluster_data <- c %>%
  mutate(derogatory = ifelse(count_tradelines_condition_derogatory==0, "None", 
                             ifelse(count_tradelines_condition_derogatory==1 |count_tradelines_condition_derogatory==2  , "Some", "Many")))%>%
  dplyr::select(derogatory,is_homeowner,credit_score_continuous, age_continuous, view_cc_details_count)%>%
  na.omit()
#Take a sample of 2000 observations to create our models
cluster_data <- cluster_data%>%
  #scale(.)%>%
  slice(1:2000)%>%
  na.omit()

#Hierarchial Modeling
#Create the distance matrix using euclidean length
d <- dist(cluster_data[,2:5], method = "euclidean")
# Hierarchical clustering using Ward's method
res.hc <- hclust(d)
grp <- cutree(res.hc, k = 3)
#table(grp, cluster_data$derogatory)
# Visualize Our hierachical model with 3 groups
plot(res.hc, cex = 0.6) # plot tree
rect.hclust(res.hc, k = 3, border = 2:5) # add rectangle

```

K-means Visualization 
===
```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.width=4, fig.height=3,fig.align='center'}
#Find optimal number of clusters
#K-means with 6 groups (as per the diagram above)

par(mfrow=c(1,2))
km.res <- kmeans(cluster_data[,2:5], 3, nstart = 25)
# Visualize our k-means
fviz_cluster(km.res, data = cluster_data[,2:5], frame.type = "convex") +
  theme_minimal()


```

K-means Visualization 
===
```{r,warning=FALSE, message=FALSE, echo=FALSE, fig.width=4, fig.height=3,fig.align='center'}
#Visualize k means
km.res <- kmeans(cluster_data[,2:4], 3, nstart = 25)
fviz_cluster(km.res, data = cluster_data[,2:4], frame.type = "convex")+
  theme_minimal()

```

Problems with Clustering Model
===
- The variables we chose were arbitrary and the hierarchical model was too naive. 
- It wasn't a good method to use with the rest of our analysis because it didn't tell us anything substantial. 
- We did not split into training/testing sets and use Cross Validation to check the model.
- We arbitrarily cut our trees at an unmotivated point.


